\documentclass[12pt, letterpaper]{article}
\usepackage{graphicx}
\title{APD Statistical Validation and Documentation}
\author{Allison Kunerth}
\date{(TODAY)}
\begin{document}
\maketitle

Model Validation is being completed by Juan,Daniel and Allison. This document serves to track this process as much as possible given the time constraints.

\textbf{Feature Selection} 
* 2.2: Analysis encompasses 6 fiscal years of data. Depending on when the end of the FY is, need to verify whether it truly is a full 6 years.
*2.3.1: Exlusions seem reasonable, including those with zero variance since they would have no impact on the outcome--and is better than excluding just low variance. 
    - MidYear QSI = 957 (0.5%)
    - Late Entry = 43,102 (20.2%)
    - Early Exit = 14,873 (7.0%)
    - No Costs = 109 (0.1%)
    - Negative Costs = 0 (0.0%)
    - Insufficient Service = 81,901 (38.5%)
    - No QSI = 128,791 (60.5%)
- Need to double check that the min, max and mean number of records is correct


Q: Do we need to include the cost requirement greater than 0 since no cases meet that definition?
A: That weas filtered from data preparation for the cache. Look at sql transformation 14 0r 13

* Cardinality is number of levels of each variable being considered. Need to check whether number is correct?


\textbf{Model 1:} 
TBD


\textbf{Model 2:}
TBD


\textbf{Model 3:}
TBD


\textbf {Model 4:} 
TBD


\textbf{Model 5:} 
TBD


\textbf{Model 6:} 
TBD


\textbf{Model 7:}
TBD


\textbf{Model 8:} 
TBD


\textbf{Model 9:} 
TBD


\textbf{Model 10:} 
TBD

    `'

\end{document}